Best Practices for Developing Monte Carlo Methodologies in Molecular Simulations [Article v\versionnumber]

### 2025-06-05 paper/paper.tex
updated orientation figure / code

The position of each particle is represented by a single vector coordinate $\mathbf{r}$, and all internal and/or orientational coordinates of the particle are represented by a single vector $\boldsymbol{\omega}$.
A particle may be composed of multiple interaction sites that are mutually bound, flexibly or rigidly or both, and the site-site interactions could be orientationally-dependent (e.g., point dipoles).
The orientation of a particle might be given implicitly (not part of $\boldsymbol{\omega}$) via the Cartesian coordinates of its interaction sites.
Alternatively, it may be specified explicitly via an appropriate orientation vector, with the location of the interaction sites (if a multi-site particle) constructed on-the-fly from the position, orientation, and internal coordinates.
In a 3-dimensional space, the orientation coordinate for a cylindrically-symmetric object requires just two independent values, e.g., a unit  vector describing the cylindrical axis direction.
For particles with no such symmetry, three independent values are needed to specify the orientation.
Euler angles ($\phi\theta\psi$) are sometimes invoked for this, but they are not particularly useful in the context of molecular simulation.
A better choice is a quaternion ${\bf q}$, which is a normalized, four-element vector.\cite{?}

@@

$10^6$ trials, $10^2$ bins, $10$ simulations, $\delta \theta_{max}=\pi/10$.

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:orientation_trials},
caption={Caption.}
]{../codes/orientation_trials.py}
\end{figure}

\begin{figure}
\begin{centering}
\includegraphics[width=8.5cm]{../figures/orientation_trials.pdf}
\caption{
\label{fig:orientation_trials}
}
\end{centering}
\end{figure}

Orientations are typically fixed during particle displacements so as not to introduce a second tunable parameter for orientation moves that cannot be decoupled from the displacement parameter during optimization.
Instead, the orientation of a single particle may be perturbed in a separate trial move.
A robust and intuitive way to accomplish this is via the axis-angle method: (1) a particle is randomly chosen among $N_m$ mobile particles; (2) a rotation axis $\hat {\bf u}$ is generated by sampling uniformly on the surface of a sphere (Sec.~\ref{sec:unit_sphere}); (3) a rotation angle $\theta$ is generated uniformly on $\pm\delta/2$ (with $\delta$ a tunable parameter).
With the rotation so defined, a rotation matrix can be computed
\begin{equation}
\mathbf{R} =
\begin{bmatrix}
C + (1 - C) u_x^2 & (1 - C) u_x u_y - S u_z & (1 - C) u_x u_z + S u_y \\
(1 - C) u_y u_x + S u_z & C + (1 - C) u_y^2 & (1 - C) u_y u_z - S u_x \\
(1 - C) u_z u_x - S u_y & (1 - C) u_z u_y + S u_x & C + (1 - C) u_z^2
\end{bmatrix}
\end{equation}
where $C =\cos\theta$ and $S = \sin\theta$.
The new coordinate of interaction site $j$ on particle $i$ at ${\bf r}_i$ is given by ${{\bf r}_{ij,{\rm n}}={\bf r}_{i}}+{\mathbf R}({\bf r}_{ij,{\rm o}}-{\bf r}_i)$.
If ${\bf r}_i$ is chosen as the centroid of the particle, then the centroid will be unchanged by the rotation (complementing the translation move, which leaves the orientation unchanged).
However, using the centroid for ${\bf r}_i$ is not a requirement; any well-defined center for the rotation may be employed (e.g., the position of a specific interaction site).

@@

Regardless of how the selected rotation is implemented, the transition probabilities are given in terms of the axis-angle selection process outlined above.
The elements of the transition probabilities for the particle orientation trial are summarized in Table~\ref{tab:lhs_rotation}.
Together they yield the forward and reverse probabilities, which in this case are again equal
\begin{equation}
\alpha_{o\rightarrow n}=\alpha_{n\rightarrow o} = \frac{\diff\mathbf{\omega}\diff\theta}{N_m 4\pi\delta}.
\label{eq:lhs_rhs_rot_fwdrev}
\end{equation}

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|c|}
 \hline
 \thead{Forward} & \thead{$\alpha_{o\rightarrow n}$} \\ [0.5ex]
 \hline
 \makecell{Choose from $N_m$} & \makecell{$1/N_m$} \\
 \hline
 \makecell{Choose rotation axis} & \makecell{$\diff\boldsymbol{\omega}/4\pi$} \\
 \hline
 \makecell{Choose rotation angle} & \makecell{$\diff\theta/\delta$} \\
 \hline\hline
 \thead{Reverse} & \thead{$\alpha_{n\rightarrow o}$}\\ [0.5ex]
 \hline
 \makecell{Choose from $N_m$} & \makecell{$1/N_m$} \\
 \hline
 \makecell{Choose rotation axis} & \makecell{$\diff\boldsymbol{\omega}/4\pi$} \\
 \hline
 \makecell{Choose rotation angle} & \makecell{$\diff\theta/\delta$} \\
 \hline
\end{tabular}
\caption{Orientation-change transition probabilities.}
\label{tab:lhs_rotation}
\end{center}
\end{table}

@@

A larger tunable parameter, $\delta$, leads to a larger rotation, and thus $\delta$ may be tuned in the same manner as the displacement parameter to reach a target acceptance for each species.
To avoid violating detailed balance, this tunable parameter should remain constant during the production simulations for the same reason as shown in Fig.~\ref{fig:lhs_nvt}.

%If desired, generation of orientations from scratch, without perturbing on an existing orientation, is accomplished by random generation of a quaternion \cite{vesely_angular_1982}.
%An example of randomly picking quaternions, and thus orientations, is shown in Section~\ref{sec:unit_sphere}.

@@

\begin{figure}
\begin{centering}
\includegraphics[width=6cm]{../figures/random_orientation.png}
\caption{
  Example output from Code Block~\ref{sim:random_orientation} shows $10^3$ uniformly random unit vectors emanating from the origin and $10^3$ unit vectors in orange randomly generated about the upwards $z$-axis (shown in black) with axis-angle rotations of a maximum angle parameter, $\delta \theta_{max}=\pi/2=90$ degrees.
\label{fig:random_orientation}
}
\end{centering}
\end{figure}

>>>>

The position of each particle is represented by a single vector coordinate $\mathbf{r}$, and all internal and/or orientational coordinates of the particle are represented by a single vector $\boldsymbol{\omega}$.
A particle may be composed of multiple interaction sites that are mutually bound, flexibly or rigidly or both, and the site-site interactions could be orientationally-dependent (e.g., point dipoles).
The orientation of a particle might be given implicitly (not part of $\boldsymbol{\omega}$) via the Cartesian coordinates of its interaction sites.
Alternatively, it may be specified explicitly via an appropriate orientation vector, with the location of the interaction sites (if a multi-site particle) constructed on-the-fly from the position, orientation, and internal coordinates.
In a 3-dimensional space, the orientation coordinate for a cylindrically-symmetric object requires just two independent values, e.g., a unit  vector describing the cylindrical axis direction.
For particles with no such symmetry, three independent values are needed to specify the orientation.
Euler angles ($\phi\theta\psi$) are sometimes invoked for this, but they are not particularly useful in the context of molecular simulation.
A better choice is a quaternion ${\bf q}$, which is a normalized, four-element vector \cite{vesely_angular_1982, karney_quaternions_2007}.

@@

Orientations are typically fixed during particle displacements so as not to introduce a second tunable parameter for orientation moves that cannot be decoupled from the displacement parameter during optimization.
Instead, the orientation of a single particle may be perturbed in a separate trial move.
A robust and intuitive way to accomplish this is via the axis-angle method: (1) a particle is randomly chosen among $N_m$ mobile particles; (2) a rotation axis $\hat {\bf u}$ is generated by sampling uniformly on the surface of a sphere (Sec.~\ref{sec:unit_sphere}); (3) a rotation angle $\theta$ is generated uniformly on $\pm\delta\theta_{max}/2$ (with $\delta\theta_{max}$ a tunable parameter).
With the rotation so defined, a rotation matrix can be computed
\begin{equation}
\mathbf{R} =
\begin{bmatrix}
C + (1 - C) u_x^2 & (1 - C) u_x u_y - S u_z & (1 - C) u_x u_z + S u_y \\
(1 - C) u_y u_x + S u_z & C + (1 - C) u_y^2 & (1 - C) u_y u_z - S u_x \\
(1 - C) u_z u_x - S u_y & (1 - C) u_z u_y + S u_x & C + (1 - C) u_z^2
\end{bmatrix}
\label{eq:rotation_matrix}
\end{equation}
where $C =\cos\theta$ and $S = \sin\theta$.
The new coordinate of interaction site $j$ on particle $i$ at ${\bf r}_i$ is given by ${{\bf r}_{ij,{\rm n}}={\bf r}_{i}}+{\mathbf R}({\bf r}_{ij,{\rm o}}-{\bf r}_i)$.
If ${\bf r}_i$ is chosen as the centroid of the particle, then the centroid will be unchanged by the rotation (complementing the translation move, which leaves the orientation unchanged).
However, using the centroid for ${\bf r}_i$ is not a requirement; any well-defined center for the rotation may be employed (e.g., the position of a specific interaction site).

@@

Regardless of how the selected rotation is implemented, the transition probabilities are given in terms of the axis-angle selection process outlined above.
The elements of the transition probabilities for the particle orientation trial are summarized in Table~\ref{tab:lhs_rotation}.
Together they yield the forward and reverse probabilities, which in this case are again equal
\begin{equation}
  \alpha_{o\rightarrow n}=\alpha_{n\rightarrow o} = \frac{\diff\mathbf{\omega}\diff\theta}{N_m 4\pi\delta\theta_{max}}.
\label{eq:lhs_rhs_rot_fwdrev}
\end{equation}

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|c|}
 \hline
 \thead{Forward} & \thead{$\alpha_{o\rightarrow n}$} \\ [0.5ex]
 \hline
 \makecell{Choose from $N_m$} & \makecell{$1/N_m$} \\
 \hline
 \makecell{Choose rotation axis} & \makecell{$\diff\boldsymbol{\omega}/4\pi$} \\
 \hline
  \makecell{Choose rotation angle} & \makecell{$\diff\theta/\delta\theta_{max}$} \\
 \hline\hline
 \thead{Reverse} & \thead{$\alpha_{n\rightarrow o}$}\\ [0.5ex]
 \hline
 \makecell{Choose from $N_m$} & \makecell{$1/N_m$} \\
 \hline
 \makecell{Choose rotation axis} & \makecell{$\diff\boldsymbol{\omega}/4\pi$} \\
 \hline
  \makecell{Choose rotation angle} & \makecell{$\diff\theta/\delta\theta_{max}$} \\
 \hline
\end{tabular}
\caption{Orientation-change transition probabilities.}
\label{tab:lhs_rotation}
\end{center}
\end{table}

@@

A larger tunable parameter, $\delta\theta_{max}$, leads to a larger rotation, and thus $\delta\theta_{max}$ may be tuned in the same manner as the displacement parameter to reach a target acceptance for each species.
To avoid violating detailed balance, this tunable parameter should remain constant during the production simulations for the same reason as shown in Fig.~\ref{fig:lhs_nvt}.

Code Block~\ref{sim:orientation_trials} tests Eqs.~\ref{eq:rotation_matrix} and \ref{eq:lhs_rhs_rot_fwdrev} with the simulation of the orientation of a single rigid body in $3$ dimensions.
Because there is no energy of interaction, the expected orientation distribution is uniform on the surface of a sphere.
In spherical coordinates, this is given by uniform distribution of the azimuthal angle $\in [-\pi,\pi]$, which is defined as the angle between the x-axis and the projection of the vector on the x-y plane, and a sine distribution for the polar angle $\in [0,\pi]$, which is the angle between the z-axis and the vector.
These analytical expectations are compared against the MC simulation in Fig.~\ref{fig:orientation_trials}.

\begin{figure}
\lstinputlisting[
language=Python,
label={sim:orientation_trials},
  caption={MC simulation of a single rigid body in three dimensions using Python 3 using $5\times 10^5$ random displacements with $\delta\theta_{max}=\pi/10$. Angle distributions were computed over $300$ equally-spaced bins with $20$ independent simulations (totaling $10^7$ orientations) to compare against theoretical expectations with approximately 68 \% of the distribution within one standard deviation.}
]{../codes/orientation_trials.py}
\end{figure}

\begin{figure}
\begin{centering}
\includegraphics[width=8.5cm]{../figures/orientation_trials.pdf}
\caption{
  The (blue) azimuthal and (red) polar probability distributions from the MC simulations described in Code Block~\ref{sim:orientation_trials} with error bars from the standard deviation of the mean of $10$ independent simulations, and theoretical expressions shown by the black lines.
\label{fig:orientation_trials}
}
\end{centering}
\end{figure}
%If desired, generation of orientations from scratch, without perturbing on an existing orientation, is accomplished by random generation of a quaternion \cite{vesely_angular_1982}.
%An example of randomly picking quaternions, and thus orientations, is shown in Section~\ref{sec:unit_sphere}.

@@

\begin{figure}
\begin{centering}
\includegraphics[width=6cm]{../figures/random_orientation.png}
\caption{
  Example output from Code Block~\ref{sim:random_orientation} shows $10^3$ uniformly random unit vectors emanating from the origin and $10^3$ unit vectors in orange randomly generated about the upwards $z$-axis (shown in black) with axis-angle rotations of a maximum angle parameter, $\delta \theta_{max}=\pi/2=90$ degrees (e.g., the orange vectors are confined to the top hemisphere).
\label{fig:random_orientation}
}
\end{centering}
\end{figure}
